---
title: "rna_differential_expression_analysis"
author: "Samantha Gallegos Perez"
date: "2026-04-26"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r packages}

library(tidyverse)
library(DESeq2)
library(ggplot2)
library(dplyr)
```

# Differential Expression Analysis Part 1: Analyzing the raw counts using DESeq2



## data upload 

```{r data upload}
#uploaded the raw counts for rna_counts 
rna_raw_counts <- read.table("rna_counts.txt", header=TRUE)
```

## data organization

The data frame includes information about the chromosome, the length, the Start and End, the strand direction and the two controls, control rep 3 at 37 degrees Celsius (37D3) and control rep 2 at 37 degrees Celsius (37D2), elevated temperature rep 2 at 46 degrees Celsius (46D2) and elevated temperature rep 3 at 46 degrees Celsius (46D3).

```{r data wrangling}

#selected only GeneID, and the replicates including: 2 controls and 2 treatments
rna_counts <- select(rna_raw_counts,"Geneid", "X37D3_control3_sorted.bam":"X46D3_elevtemp3_sorted.bam")

#renaming the columns from .bam files to descriptive features
colnames(rna_counts)[2] ="Control_rep_37C_3"
colnames(rna_counts)[3] ="Control_rep_37C_2"
colnames(rna_counts)[4] ="Elevtemp_rep_46C_2"
colnames(rna_counts)[5] ="Elevtemp_rep_46C_3"

#make the Geneid the row id 
rna_counts <- rna_counts %>% remove_rownames %>% column_to_rownames(var="Geneid")

```

## Metadata Creation

```{r metadata}
#creating a metadata based on our replicates

feature_names <- c(colnames(rna_counts))
#specify the conditions for each sample

#In my sample names, 37C replicates are controls, 46C samples are treated replicates

condition <- c("control", "control", "treated", "treated")

# generate the metadata data frame
meta_data <- data.frame(feature_names, condition)

# make the sample name the row id
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="feature_names")

meta_data
rna_counts

#check to see if the metadata rownames and rna_counts columns match: TRUE for both
all(colnames(rna_counts) %in% rownames(meta_data)) 
all(colnames(rna_counts) == rownames(meta_data))
```

## Differential Analysis

Create a DESeq object for analyses 
```{r DESeq obj}
#create DESeq object using the metadata created, the rna_counts data frame and the design by condition
d_obj <- DESeqDataSetFromMatrix(countData = rna_counts,
                              colData = meta_data,
                              design = ~ condition)

#converting variables with char data types into factors as the object is created.

#looking at the obj, there should be a dim of 106 (genes) and 4 (replicates)
d_obj


#pre-filtering low counts out of the analysis

#this was used to increase speed of transformation and improve visualizations

#filtering out counts less than 10
#keep <- rowSums(counts(d_obj)) >= 10
#d_obj <- d_obj[keep,]

#removed 61 features that had less than 10 counts
d_obj

#setting the reference factor to control manually
d_obj$condition <- relevel(d_obj$condition, ref = 'control')

#getting normalized counts 
d_obj <- estimateSizeFactors(d_obj)
normalized_counts <- counts(d_obj, normalized = TRUE)
```


## Run Differential Expression


```{r differential analysis}

#using DESeq() to extract results table with log2 fold changes, p values and adjusted p values
d_obj <- DESeq(d_obj)
res <- results(d_obj)
res

#to output the summary of the table 
summary(res)

#only one sample 
sum(res$padj < 0.1, na.rm=TRUE)
```


## Visualizing Results 

```{r visualizations}
#creating the res results into a dataframe
data <- data.frame(res)
head(data)

#creating a PCA plot
rld <- rlog(d_obj)
plotPCA(rld)


```


```{r volcano plot}



#volcano plot

data <- data %>%
  mutate(
  Expression = case_when(log2FoldChange >= log(1) & padj <= 0.05 ~ "Up-regulated",
  log2FoldChange <= -log(1) & padj <= 0.05 ~ "Down-regulated",
  TRUE ~ "Unchanged")
  )
head(data)



#getting the top 10
top <- 10
# we are getting the top 10 up and down regulated genes by filtering the column Up-regulated and Down-regulated and sorting by the adjusted p-value. 
top_genes <- bind_rows(
  data %>%
  filter(Expression == 'Up-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top),
  data %>%
  filter(Expression == 'Down-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top)
  )
# create a datframe just holding the top 10 genes
Top_Hits = head(arrange(data,pvalue),10)
Top_Hits


data$label = if_else(rownames(data) %in% rownames(Top_Hits), rownames(data), "")
# basic plot
p1 <- ggplot(data, aes(log2FoldChange, -log(pvalue,10))) + # -log10 conversion
geom_point( size = 2/5) +
xlab(expression("log"[2]*"FC")) +
ylab(expression("-log"[10]*"P-Value")) +
xlim(-4.5, 4.5)
p1


#basic plot with line + red for p < 0.05
p2 <- ggplot(data, aes(log2FoldChange, -log(pvalue,10))) + # -log10 conversion
geom_point(aes(color = Expression), size = 2/5) +
#geom_hline(yintercept= -log10(0.05), linetype="dashed", linewidth = .4) +
xlab(expression("log"[2]*"FC")) +
ylab(expression("-log"[10]*"P-Value")) +
scale_color_manual(values = c("firebrick3", "black", "firebrick3")) +
xlim(-4.5, 4.5) +
theme(legend.position = "none")
p2



```


# Differential Expression Analysis Part 2: Analyzing gene expression using EdgeR

