---
title: "rna_differential_expression_analysis"
author: "Samantha Gallegos Perez"
date: "2026-04-26"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r packages}

library(tidyverse)
library(DESeq2)
library(ggplot2)
library(dplyr)
library(limma)
library(Glimma)
library(edgeR)
#associated annotation data for pseudomonas arguenosas from Bioconductor
#install.packages("refseqR")
library(biomaRt)
```

# Differential Expression Analysis Part 1: Analyzing the raw counts using DESeq2



## data upload 

```{r data upload}
#uploaded the raw counts for rna_counts 
rna_raw_counts <- read.table("rna_counts.txt", header=TRUE)
```

## data organization

The data frame includes information about the chromosome, the length, the Start and End, the strand direction and the two controls, control rep 3 at 37 degrees Celsius (37D3) and control rep 2 at 37 degrees Celsius (37D2), elevated temperature rep 2 at 46 degrees Celsius (46D2) and elevated temperature rep 3 at 46 degrees Celsius (46D3).

```{r data wrangling}

#selected only GeneID, and the replicates including: 2 controls and 2 treatments
rna_counts <- select(rna_raw_counts,"Geneid", "X37D3_control3_sorted.bam":"X46D3_elevtemp3_sorted.bam")

#renaming the columns from .bam files to descriptive features
colnames(rna_counts)[2] ="Control_rep_37C_3"
colnames(rna_counts)[3] ="Control_rep_37C_2"
colnames(rna_counts)[4] ="Elevtemp_rep_46C_2"
colnames(rna_counts)[5] ="Elevtemp_rep_46C_3"

#make the Geneid the row id 
rna_counts <- rna_counts %>% remove_rownames %>% column_to_rownames(var="Geneid")

```

## Metadata Creation

```{r metadata}
#creating a metadata based on our replicates

feature_names <- c(colnames(rna_counts))
#specify the conditions for each sample

#In my sample names, 37C replicates are controls, 46C samples are treated replicates

condition <- c("control", "control", "treated", "treated")

# generate the metadata data frame
meta_data <- data.frame(feature_names, condition)

# make the sample name the row id
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="feature_names")

meta_data
rna_counts

#check to see if the metadata rownames and rna_counts columns match: TRUE for both
all(colnames(rna_counts) %in% rownames(meta_data)) 
all(colnames(rna_counts) == rownames(meta_data))
```

## Differential Analysis

Create a DESeq object for analyses 
```{r DESeq obj}
#create DESeq object using the metadata created, the rna_counts data frame and the design by condition
d_obj <- DESeqDataSetFromMatrix(countData = rna_counts,
                              colData = meta_data,
                              design = ~ condition)

#converting variables with char data types into factors as the object is created.

#looking at the obj, there should be a dim of 106 (genes) and 4 (replicates)
d_obj


#pre-filtering low counts out of the analysis

#this was used to increase speed of transformation and improve visualizations

#filtering out counts less than 10
#keep <- rowSums(counts(d_obj)) >= 10
#d_obj <- d_obj[keep,]

#removed 61 features that had less than 10 counts
d_obj

#setting the reference factor to control manually
d_obj$condition <- relevel(d_obj$condition, ref = 'control')

#getting normalized counts 
d_obj <- estimateSizeFactors(d_obj)
normalized_counts <- counts(d_obj, normalized = TRUE)
```


## Run Differential Expression


```{r differential analysis}

#using DESeq() to extract results table with log2 fold changes, p values and adjusted p values
d_obj <- DESeq(d_obj)
res <- results(d_obj)
res

#to output the summary of the table 
summary(res)

#only one sample 
sum(res$padj < 0.1, na.rm=TRUE)
```


## Visualizing Results 

```{r visualizations}
#creating the res results into a dataframe
data <- data.frame(res)
head(data)

#creating a PCA plot
rld <- rlog(d_obj)
plotPCA(rld)


```


```{r volcano plot}



#volcano plot

data <- data %>%
  mutate(
  Expression = case_when(log2FoldChange >= log(1) & padj <= 0.05 ~ "Up-regulated",
  log2FoldChange <= -log(1) & padj <= 0.05 ~ "Down-regulated",
  TRUE ~ "Unchanged")
  )
head(data)



#getting the top 10
top <- 10
# we are getting the top 10 up and down regulated genes by filtering the column Up-regulated and Down-regulated and sorting by the adjusted p-value. 
top_genes <- bind_rows(
  data %>%
  filter(Expression == 'Up-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top),
  data %>%
  filter(Expression == 'Down-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top)
  )
# create a datframe just holding the top 10 genes
Top_Hits = head(arrange(data,pvalue),10)
Top_Hits


data$label = if_else(rownames(data) %in% rownames(Top_Hits), rownames(data), "")
# basic plot
p1 <- ggplot(data, aes(log2FoldChange, -log(pvalue,10))) + # -log10 conversion
geom_point( size = 2/5) +
xlab(expression("log"[2]*"FC")) +
ylab(expression("-log"[10]*"P-Value")) +
xlim(-4.5, 4.5)
p1


#basic plot with line + red for p < 0.05
p2 <- ggplot(data, aes(log2FoldChange, -log(pvalue,10))) + # -log10 conversion
geom_point(aes(color = Expression), size = 2/5) +
#geom_hline(yintercept= -log10(0.05), linetype="dashed", linewidth = .4) +
xlab(expression("log"[2]*"FC")) +
ylab(expression("-log"[10]*"P-Value")) +
scale_color_manual(values = c("firebrick3", "black", "firebrick3")) +
xlim(-4.5, 4.5) +
theme(legend.position = "none")
p2



```


# Differential Expression Analysis Part 2: Analyzing gene expression using EdgeR

## Uploading the data 

```{r uploading the data}
#upload the data 
rna_data <- read.table("RNA_data.txt",header=TRUE)
```

The data frame includes information about the chromosome, the length, the Start and End, the strand direction, three controls, all at temperature 37C (36D1, 37D2, 37D3), and three treatments with elevated temperature at 46C (46D1, 46D2, 46D3).


```{r maybe delete?}
#currently remove unecessary columns for now 
rna_data2 <- rna_data[c(-2:-6)]

#change the row indices to the geneid names
rna_data2 <- rna_data2 %>% remove_rownames %>% column_to_rownames(var="Geneid")

#create a column for gene lengths?
#rna_data2["gene_length"] <- rna_data["End"] - rna_data["Start"]

```



```{r create DGE object}
x <- DGEList(rna_data2)
x
class(x)
#there are 106 genes and 6 features: 3 controls, 3 treatments
dim(x)
```

## organising sample information
```{r creating metadata}
#created column names by substring to temp and relicate numbers
samplenames <- substring(colnames(x), 2, 5)
samplenames

colnames(x) <- samplenames
group <- as.factor(c("control", "control", "control", "heat_exp_treatment",
                     "heat_exp_treatment", "heat_exp_treatment"))
x
x$samples$group <- group
x$samples
x


```


Created another dataframe for annotated gene information

```{r genes dataframe}
geneid <- rownames(x)
geneid

# 2. Read the GFF/GTF file (downloaded from Pseudomonas.com)
gff <- import("genomic.gff")

# 3. Convert to data frame and filter for gene names
gff_df <- as.data.frame(gff)
# Assuming IDs are in 'gene_id' or 'Name' column and Type is 'gene'
gene_map <- gff_df[gff_df$type == "gene",]

#change name from Name to geneid to merge with my_genes column 
names(gene_map)[13] <- "geneid"
# 4. Merge with your gene list
my_genes <- data.frame(geneid)
genes <- merge(gene_map, my_genes, by = "geneid")

x$genes <- genes
x
```


```{r transformations}



```







