---
title: "rna_differential_expression_analysis"
author: "Samantha Gallegos Perez"
date: "2026-04-26"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r packages}
library(dplyr)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(dplyr)
library(limma)
library(Glimma)
library(edgeR)
#associated annotation data for pseudomonas arguenosas from Bioconductor
#install.packages("refseqR")
library(biomaRt)
library(rtracklayer)
library(EnhancedVolcano)
library(pheatmap)
```

# Differential Expression Analysis Part 1: Analyzing the raw counts using DESeq2

## data upload 

```{r data upload}
#uploaded the raw counts for rna_counts 
rna_raw_counts <- read.table("rna_data.txt", header=TRUE)
```

## data organization

The data frame includes information about the chromosome, the length, the Start and End, the strand direction and the two controls, control rep 3 at 37 degrees Celsius (37D3) and control rep 2 at 37 degrees Celsius (37D2), elevated temperature rep 2 at 46 degrees Celsius (46D2) and elevated temperature rep 3 at 46 degrees Celsius (46D3).

```{r data wrangling}

#selected only GeneID, and the replicates including: 2 controls and 2 treatments
rna_counts <- select(rna_raw_counts,"Geneid", 7:12)
lengths <- rna_raw_counts[6]

#renaming the columns from .bam files to descriptive features
colnames(rna_counts)[2] ="Control_rep_1"
colnames(rna_counts)[3] ="Control_rep_2"
colnames(rna_counts)[4] ="Control_rep_3"
colnames(rna_counts)[5] ="Elevtemp_rep_1"
colnames(rna_counts)[6] ="Elevtemp_rep_2"
colnames(rna_counts)[7] ="Elevtemp_rep_3"

#make the Geneid the row id 
rna_counts <- rna_counts %>% remove_rownames %>% column_to_rownames(var="Geneid")

```

## Metadata Creation

```{r metadata}
#creating a metadata based on our replicates

feature_names <- c(colnames(rna_counts))
#specify the conditions for each sample

#In my sample names, 37C replicates are controls, 46C samples are treated replicates

condition <- c("control", "control","control", "heat-shock", "heat-shock", "heat-shock")

# generate the metadata data frame
meta_data <- data.frame(feature_names, condition)

# make the sample name the row id
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="feature_names")

meta_data
rna_counts

#check to see if the metadata rownames and rna_counts columns match: TRUE for both
all(colnames(rna_counts) %in% rownames(meta_data)) 
all(colnames(rna_counts) == rownames(meta_data))
```

## Differential Analysis

Create a DESeq object for analyses 
```{r DESeq obj}
#create DESeq object using the metadata created, the rna_counts data frame and the design by condition
d_obj <- DESeqDataSetFromMatrix(countData = rna_counts,
                              colData = meta_data,
                              design = ~ condition)

#converting variables with char data types into factors as the object is created.

#looking at the obj, there should be a dim of 106 (genes) and 4 (replicates)
d_obj


#pre-filtering low counts out of the analysis

#this was used to increase speed of transformation and improve visualizations

#filtering out counts less than 10
keep <- rowSums(counts(d_obj)) >= 10
d_obj <- d_obj[keep,]

#removed 61 features that had less than 10 counts
d_obj

#setting the reference factor to control manually
d_obj$condition <- relevel(d_obj$condition, ref = 'control')

#getting normalized counts 
d_obj <- estimateSizeFactors(d_obj)
normalized_counts <- counts(d_obj,normalized= TRUE)



```


## Run Differential Expression


```{r differential analysis}

#using DESeq() to extract results table with log2 fold changes, p values and adjusted p values
d_obj <- DESeq(d_obj)

res <- results(d_obj, alpha=0.05)
res

#to output the summary of the table 
summary(res)



```


## Visualizing Results 

```{r visualizations}
#creating the res results into a dataframe
data <- data.frame(res)
head(data)

#creating a PCA plot
rld <- rlog(d_obj)

png("pca_biplot.png", width=800,height=600)
plotPCA(rld) + labs(title = "PCA", size = 80) +
  geom_point(size = 6, aes(shape = group)) + 
  scale_color_manual(values = c("lightblue", "purple")) +
  stat_ellipse(type = "norm", level = 0.95) +
  theme(plot.title = element_text(hjust = 0.5))
dev.off()


```


```{r volcano plot}

#volcano plot

data <- data %>%
  mutate( 
  Expression = case_when(log2FoldChange >= 1 & padj <= 0.1 ~ "Up-regulated",
  log2FoldChange <= -1 & padj <= 0.1 ~ "Down-regulated",
  TRUE ~ "Unchanged")
  )
head(data)

data$gene_names = rownames(data)

```

```{r}


#getting the top 10
top <- 10
# we are getting the top 10 up and down regulated genes by filtering the column Up-regulated and Down-regulated and sorting by the adjusted p-value. 
top_genes <- bind_rows(
  data %>%
  filter(Expression == 'Up-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top),
  data %>%
  filter(Expression == 'Down-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top)
  )
# create a datframe just holding the top 10 genes
Top_Hits = head(arrange(data,pvalue),10)
Top_Hits
```


```{r heatmap}
rld <- vst(d_obj, blind=FALSE)
top_genes <- head(order(res$padj), 50)
mat <- assay(rld)[top_genes, ]
mat <- mat - rowMeans(mat)

png("heatmap.png", width=800,height=600)
pheatmap(mat, 
         annotation_col = meta_data, 
         main = "Top 20 Variable Genes: Heat-Shock vs Control",
         color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
         border_color = NA)
dev.off()



```

```{r volcano plot}
png("volcanoplot.png", width=1000,height=800)
EnhancedVolcano(data,
    lab = rownames(data),
    x = 'log2FoldChange',
    y = 'padj',                # Usually better to use padj than raw p-value
    title = 'Heat-Shock vs Control',
    pCutoff = 0.05,            # Significance threshold
    FCcutoff = 1.0,            # Fold change threshold (log2 scale)
    pointSize = 3.0,
    labSize = 4.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2'),
    legendPosition = 'right')
dev.off()
```






