---
title: "rna_differential_expression_analysis"
author: "Samantha Gallegos Perez"
date: "2026-04-26"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r packages}

library(tidyverse)
library(DESeq2)
library(ggplot2)
library(dplyr)
```

# Differential Expression Analysis

## data upload 

```{r data upload}
#uploaded the raw counts for rna_counts 
rna_raw_counts <- read.table("rna_counts.txt", header=TRUE)
```

## data organization

The data frame includes information about the chromosome, the length, the Start and End, the strand direction and the two controls, control rep 3 at 37 degrees Celsius (37D3) and control rep 2 at 37 degrees Celsius (37D2), elevated temperature rep 2 at 46 degrees Celsius (46D2) and elevated temperature rep 3 at 46 degrees Celsius (46D3).

```{r data wrangling}

#selected only GeneID, and the replicates including: 2 controls and 2 treatments
rna_counts <- select(rna_raw_counts,"Geneid", "X37D3_control3_sorted.bam":"X46D3_elevtemp3_sorted.bam")

#renaming the columns from .bam files to descriptive features
colnames(rna_counts)[2] ="Control_rep_37C_3"
colnames(rna_counts)[3] ="Control_rep_37C_2"
colnames(rna_counts)[4] ="Elevtemp_rep_46C_2"
colnames(rna_counts)[5] ="Elevtemp_rep_46C_3"

```

## Metadata Creation

```{r metadata}




```



